Time Stretching of the GeV Emission of GRBs: Fermi LAT data vs geometrical model

### 2014-01-21 Paper/gammaRays.tex
Revised the Introduction

Gamma Ray Bursts (GRBs) are among the most energetic events in the Universe, therefore they might provide new knowledge for particle physics. Modern observatories, specifically Fermi \cite{Ackermann:2012kna} and Swift \cite{Gehrels:2004gu} made possible to study these explosions extensively \cite{Vianello:2013ela,Gehrels:2013xd}. These and previous studies led to several interesting results. For example, the total energy emitted in gamma-rays during a burst was found to be similar among different bursts within an order of magnitude \cite{Bloom:2003wy}. This suggests that most of the bursts must have similar energetics in their rest frame. Temporal variations of spectra were also studied. The spectral lags were found between different low energy bands \cite{Yi:2005ht}. The very high energy radiation was discovered to be extended if compared to x-ray emission \cite{Vianello:2013ela,Lange:2013uh}. Moreover, the Large Area Telescope (LAT) of Fermi allows us to explore the spectral variations even between high energy bands like $(\unit[100]{MeV}, \unit[1]{GeV})$ and $(\unit[1]{GeV}, \unit[300]{GeV})$ (we'll call them low and high energy bands). In this paper we use the Kolmogorov-Smirnov test to compute the time stretching of radiation in one of these bands compared to the other.

We take the GRB list from the Fermi-LAT catalog \cite{Ackermann:2013zfa}. Tables 2 and 4 in \cite{Ackermann:2013zfa} provide the time, durations and locations of the bursts, which we use to download events data from the LAT Data Server\footnote{\url{http://fermi.gsfc.nasa.gov/cgi-bin/ssc/LAT/LATDataQuery.cgi}}. We also download events during a long time interval before the burst, and use the technique introduced in \cite{Rubtsov:2011qq} to estimate background radiation in both energy bands.
Most of the bursts in the catalog, however, do not have enough events with over $\unit[1]{GeV}$ energies to do desired computations, so we choose only those bursts for study, for which at least $10$ photons were detected in the high energy band. This leaves $3$ of them: 080916C \cite{Tajima:2009az}, 090902B \cite{Abdo:2009pg} and 090926A \cite{Bregeon:2011bu}.

@@

Finally, we stretch the high energy CDF by different factors, and compare it to the low energy CDF using the KS-test. If we require $2\sigma$-significant probability to exclude a particular stretching factor, then the stretching factor of $1$ (which means no stretching) is allowed for GRBs 080916C and 090902B. However, for 090926A, the high energy light curve should be stretched with respect to the low energy one.

We propose that this result can be explained by the curvature effects (that is the effects of the jet geometry). These effects were explored by multiple authors \cite{Nakamura:2001kd,Shen:2005ea,Shenoy:2013cba}. However, these studies were only concerned with x-ray radiation, and, more importantly, they assumed that the distribution of radiation sources is homogeneous throughout the jet. We propose a contrary idea: that the highest energy radiators are concentrated near the axis of the burst, that is the burst opening angle depends on energy. We cannot prove this assumption rigorously, but we have arguments supporting it.

First of all, there are around 750 GRBs detected by GBM, half of which were in the LAT field of view \cite{Vianello:2013ela}. However, only about 30 of them were detected by the LAT, and only 3 of them were bright in the high energy band. If we extrapolate the uniform jet model to very high energies, it means that these groups of bursts are internally different: some of them produce VHE radiation, and other do not. These differences are hard to explain given that burst energetics are similar \cite{Bloom:2003wy}. Nevertheless, these differences in burst counts can easily be explained by our model. In our model, the opening angle of low energy jet is much larger than the opening angle of a high energy jet. Therefore the most common scenario for an observable jet is that the off-axis angle of the observer is smaller than low energy jet opening angle, but larger than the opening angle of a high energy jet. Because of that, most of the bursts are only seen at low energies. The 3 bursts we study in this paper were seen from the lowest off-axis angles.

Secondly, consider the jet plasma right after emission from the central engine. There, two processes happen simultaneously:
\begin{enumerate}
	\item{Plasma particles near the jet boundary change directions, therefore increasing the jet opening angle.}
	\item{Plasma particles lose energy, therefore decreasing radiation frequency emitted from the jet.}
\end{enumerate}
We argue that these two processes are correlated, because they happen due to the same particle interaction processes. And since they are correlated, jets with large opening angles will have lower energies, which is our main assumption.

\begin{figure}[modelOverview]
        \centering
        \includegraphics[width=0.7\textwidth]{modelOverview}
        \caption{
        	Model Overview. Here the red and the blue cones represent volumes through which low and high energy plasma propagates. In the case depicted the observer's off-axis angle is smaller than the opening angle of the low energy jet, so most of the observable low energy photons will travel along the straight line from the central engine (due to the relativistic beaming effect). In the same time, the observer's off-axis angle is larger that the opening angle of the high energy jet, so high energy radiation will still originate near the center of the jet (because it is the only place where there are high energy radiators). The observation time of a photon is a sum of two things: the time interval spent in plasma as a radiator (which is approximately a distance from the central engine to the point of emission); and the time interval from emission to detection (which is the distance from the point of emission to the observer's location). Given a position of the shock front, this sum is larger for high energy photons. Because of that, high energy emission will be observed later throughout the burst duration, therefore the high energy light curve will be stretched.
        }
        \label{fig:modelOverview}
\end{figure}

\begin{figure}[sampleLightCurve]
	\centering
	\includegraphics[width=0.49\textwidth]{sampleLightCurve}
	\caption{
		High and low energy light curves produced by the geometrical model. Parameter values are the same as discussed in the section \ref{sec:parameters}. Cosmological redshift of the burst $z = 1$. Observer's off-axis angle $\chi = 0.01$.
	}
	\label{fig:sampleLightCurve}
\end{figure}

Finally, this assumption is able to explain the non-trivial stretching factor of the GRB 090926A. The detailed computations will be described in the following sections, but you can see why the stretching is happening on the figure \ref{fig:modelOverview}. The computations confirm that this qualitative picture is indeed correct, you can see a sample light curve on the figure \ref{fig:sampleLightCurve}, and predicted distribution of stretching factors on the figure \ref{fig:kappaDistribution}.

>>>>

Gamma Ray Bursts (GRBs) are among the most energetic events in the Universe, therefore they might provide new knowledge for particle physics. Modern observatories, specifically Fermi \cite{Ackermann:2012kna} and Swift \cite{Gehrels:2004gu} made possible to study these explosions extensively \cite{Vianello:2013ela,Gehrels:2013xd}. These and previous studies led to several interesting results. For example, the total energy emitted in gamma-rays during a burst was found to be similar among different bursts within an order of magnitude \cite{Bloom:2003wy}. This suggests that most of the bursts must have similar energetics in their rest frame. Temporal variations of spectra were also studied. The spectral lags were found between different low energy bands \cite{Yi:2005ht}. The very high energy radiation was discovered to be extended relative to x-ray emission \cite{Vianello:2013ela,Lange:2013uh}. Moreover, the Large Area Telescope (LAT) of Fermi allows us to explore the spectral variations even between high energy bands like $(\unit[100]{MeV}, \unit[1]{GeV})$ and $(\unit[1]{GeV}, \unit[300]{GeV})$ (we call them low and high energy bands throughout the paper). In this study we use the Kolmogorov-Smirnov test to compute the time stretching of radiation in one of these bands compared to the other.

We take the GRB list from the Fermi-LAT catalog \cite{Ackermann:2013zfa}. Tables 2 and 4 in \cite{Ackermann:2013zfa} provide the time, durations and locations of the bursts, which we use to download observational data from the LAT Data Server\footnote{\url{http://fermi.gsfc.nasa.gov/cgi-bin/ssc/LAT/LATDataQuery.cgi}}. We also download events during a day before the burst, and use the technique introduced in \cite{Rubtsov:2011qq} to estimate background radiation in both energy bands.
Most of the bursts in the catalog, however, do not have enough events with over $\unit[1]{GeV}$ energies to do desired computations, so we choose only those bursts, from which at least $10$ photons were detected in the high energy band. This leaves $3$ of them: 080916C \cite{Tajima:2009az}, 090902B \cite{Abdo:2009pg} and 090926A \cite{Bregeon:2011bu}.

@@

Finally, we stretch the high energy CDF by different factors, and compare it to the low energy CDF using the KS-test. If we require $2\sigma$-significant probability to exclude a particular stretching factor, then the stretching factor of $1$ (which means no stretching) is allowed for GRBs 080916C and 090902B. However, for 090926A, the stretching factor of $1$ is prohibited, so the high energy light curve should be stretched with respect to the low energy one.

We propose that this result can be explained by the curvature effects (that is the effects of the jet geometry). These effects were explored by multiple authors \cite{Nakamura:2001kd,Shen:2005ea,Shenoy:2013cba}. However, these studies were only concerned with x-ray radiation, and, even more importantly, they assumed that the distribution of radiation sources is homogeneous throughout the jet. We propose a contrary idea: that the highest energy radiators are concentrated near the axis of the burst, meaning that the burst opening angle depends on energy. We cannot prove this assumption rigorously, but have arguments supporting it.

First of all, there are around 750 GRBs detected by GBM, half of which were in the LAT field of view at the moment of observation \cite{Vianello:2013ela}. However, only about 30 of them were detected by the LAT, and only 3 of them were bright in the high energy band. If we extrapolate the uniform jet model to very high energies, this observation would mean that these groups of bursts are internally different: some of them produce VHE radiation, and other do not. These differences are hard to explain given that burst energetics are similar \cite{Bloom:2003wy}. Nevertheless, these differences in burst counts can easily be explained by our model. In our model, the opening angle of a jet is inversely proportional to the energy of photons it radiates. Therefore the most common scenario is that the off-axis angle of the observer is smaller than the low energy jet opening angle, but larger than the opening angle of a high energy jet. Because of that, most of the bursts can only be seen at low energies. The 3 bursts we study in this paper, were seen, according to our model, from the lowest off-axis angles.

Second, consider the plasma right after its ejection from the central engine. Two processes happen there simultaneously:
\begin{enumerate}
	\item{Particles near the jet boundary change their movement directions, therefore increasing the opening angle of the jet.}
	\item{Particles lose energy, therefore decreasing radiation frequency emitted by the jet.}
\end{enumerate}
We argue that these two processes are correlated, for they happen due to the same particle interaction processes. And since they are correlated, jets with larger opening angles should have lower energies, which is the assumption we are trying to justify.

\begin{figure}[modelOverview]
        \centering
        \includegraphics[width=0.7\textwidth]{modelOverview}
        \caption{
        	Model Overview. Here the red and the blue cones represent the volumes through which low and high energy plasma propagates. In case depicted the observer's off-axis angle is smaller than the opening angle of the low energy jet, so, due to the relativistic beaming effect, the most of the observable low energy photons will travel along the straight line from the central engine. Also, the observer's off-axis angle is larger that the opening angle of the high energy jet, so the high energy radiation will still originate near the center of the jet (because it is the only place where there are high energy radiators). The observation time of a photon is a sum of two things: the time interval spent in plasma as a radiator (which approximately equals to the distance from the central engine to the point of emission); and the time interval from emission to detection (which is the distance from the point of emission to the observer's location). Given a position of the shock front, this sum is larger for high energy photons. Because of that, high energy emission will be observed later throughout the burst duration, therefore the high energy light curve will be stretched.
        }
        \label{fig:modelOverview}
\end{figure}

\begin{figure}[sampleLightCurve]
	\centering
	\includegraphics[width=0.49\textwidth]{sampleLightCurve}
	\caption{
		High and low energy light curves produced by the geometrical model. Parameter values are the same as discussed in the section \ref{sec:parameters}. Redshift $z = 1$. Observer's off-axis angle $\chi = 0.01$.
	}
	\label{fig:sampleLightCurve}
\end{figure}

Finally, our assumption is able to explain the non-trivial stretching factor of the GRB 090926A. Detailed computations will be described in the following sections, but you can see the main idea on the figure \ref{fig:modelOverview}. The computations confirm that this qualitative picture is indeed correct. You can see a sample light curve on the figure \ref{fig:sampleLightCurve}, and predicted distribution of stretching factors on the figure \ref{fig:kappaDistribution}.
